---
title: "Quick Introduction to R"
output: html_notebook
---

This is an R notebook version of the R introduction notes.
A notebook is like a Markdown document except the code
fragments are executed and the output is included in
the notebook.

To render the notebook, use Rstudio, or install the `rmarkdown`
package in R and render from the command line:

```shell
Rscript -e "rmarkdown::render ('r-notebook.Rmd')"
```

# Help

An essential function in R is `help`.
Since the function is interactive, it is not suitable for inclusion in notebooks.

```r
help (help)
```

# Packages

A significant benefit of R is in the availability of various packages or libraries.
Typically, these are available through the [CRAN archive](https://cran.r-project.org),
but R can install them directly using the `install.packages` command. Once a package
is installed, the exported symbols are loaded using the `library` command. We will
be using the [Tidyverse Package Collection](https://www.tidyverse.org).

```{r}
library ('tidyverse')
```

# Vectors

Among the basic vector types in R are `logical`, `integer`, `double`, `complex` and `character`.
Scalars are in fact vectors of length one.
Indices start at 1.

```{r}
x <- c (2, 3, 4, 3, 3, 2, 5, 1, 2, 3)
print (x)
```

Generate 5 numbers from normal distribution with mean of 10 and SD of 2 (N (10, 4)).

```{r}
x <- rnorm (5, 10, 2)
print (x)
```

# Viewing Large Data

```{r}
x <- rnorm (1000, 10, 2)
head (x)
```

# Plotting with Plain R

## Simple Plot

```{r}
plot (x)
plot (x, type='l') # Lines instead of dots
```

## Boxplot

Extremely useful to get a quick feel for data distribution.
Can also be used for roughly comparing alternatives.

```{r}
boxplot (x)
```

## Histogram

To view the empirical distribution:

```{r}
hist (x)
```

## Zooming

```{r}
plot (x [2:20])
plot (x, xlim=c (2,20), ylim=c (3,8))
```

## Colors and Legend

```{r}
x1 <- rnorm (100, 10, 2)
x2 <- rnorm (100, 14, 2)
x3 <- rnorm (100, 18, 2)
plot (x1, ylim = c (0, max (c (x1, x2, x3))), col = 'blue')
points (x2, col = 'red')
points (x3, col = 'green')
legend ('topright', legend = c ('x1', 'x2', 'x3'), col = c ('blue', 'red', 'green'), pch = 20)
```

# Plotting with Tidyverse GGPlot

## Reading Data from File

```{r}
x <- read_csv ('https://d3s.mff.cuni.cz/files/teaching/nswi131/labs/01/data.csv.gz', col_types = 'cnii')
print (x)
```

## Plot One Run

```{r}
xx <- x |> filter (benchmark == 'alpha', run == 1)
```

Note that `xx` is again a tibble.

```{r}
ggplot (xx, aes (x = loop, y = score)) + geom_point ()
```

The `ggplot` function creates a plot object with given data as source.
The `aes` information provides defaults for data dependent values.
The `geom` functions add actual output to the plot object.

## More Runs with Colors

```{r}
xx <- x |> filter (benchmark == 'alpha')
ggplot (xx, aes (x = loop, y = score, color = run)) + geom_line (alpha = 0.2)
```

## Boxplots

Factors are category types where each category has a name.
Internally factors are represented as integers.
Beware mixing factors with different levels.

```{r}
xx <- x |> filter (benchmark == 'alpha')
ggplot (xx, aes (x = factor (run), y = score)) + geom_boxplot ()
```

## Histograms

```{r}
xx <- x |> filter (benchmark == 'alpha')
ggplot (xx, aes (x = score)) + geom_histogram (bins = 100)
ggplot (xx, aes (x = score, color = factor (run))) + geom_histogram (bins = 100)
```

## Facets

```{r}
ggplot (x, aes (x = score)) + geom_histogram (bins = 100) + facet_wrap (vars (benchmark))
ggplot (x, aes (x = score)) + geom_histogram (bins = 100) + facet_wrap (vars (benchmark), scale = 'free')
```
